Smart Model Elimination for Efficient Automated Machine Learning in Healthcare

### 2024-06-26 paper/paper.tex
new diagram
Update paper.tex

Machine learning (ML) is a type of artificial intelligence (AI) that uses searching and trend-determining to make predictions. ML algorithms are trained and tested on a given dataset to become "smarter". Once they've learned the patterns typical of that data, they can make predictions based on similar data. These algorithms also require additional processes, however, like hyperparamater tuning, and can be modified in conjunction with one another, like in ensemble learning. This makes the process complex and time-inefficient. To do this on a larger scale and at greater speeds, recent research and effort have been put into developing AutoML frameworks [1].

@@

When a user inputs a dataset into SMEML, it first extracts the attributes from the dataset and then feeds them into the SME to get the predicted rank of each model. Since we have determined that we can predict a top two model within the predicted top eight quite confidently, we can eliminate the bottom 20 models from the list. We run Bayseian optimization using scikit-optimize [16] on the top eight models to get the best hyperparamters for each model. We select the top models that have significantly better performance than other models. These models are stacked together to form the final model. This process is shown in Figure 1.
\begin{table}
  \caption{Attributes extracted from a dataset}
  \label{model-options-table}
  \centering
  \begin{tabular}{llll}
    \toprule
    Attribute & Formula & Purpose & Weight\\
    \midrule
    \# of Rows & $|\mathbf{A}|_r$ & DS & .0422 \\
    \hline
    \# of Columns & $|\mathbf{A}|_c$ & DS & .0415 \\
    \hline
    Target Distribution &
    $\frac{\max(\text{count}_1, \text{count}_0)}{|\mathbf{A}|_r}$
    & DT & .0489 \\
    \hline
    Numerical Columns &
    $\frac{|\mathbf{A}_{num}|}{|\mathbf{A}|_{c}}$
    & DT & .0428 \\
    \hline
    Binary Categorical Columns &
    $\frac{|\mathbf{A}_{bin}|}{|\mathbf{A}|_{c}}$
    & DT & .0117 \\
    \hline
    Mean \# of Distinct Values  & $\mu \left\{\text{\# of distinct values in column}\right\}$ & DT & .0076 \\ 
    \hline
    Mean IQR  & $\mu \left\{\text{IQR of column}\right\}$ & DD & .0594 \\ 
    \hline
    STD of IQR  & $\sigma \left\{\text{IQR of column}\right\}$ & DD & .0700 \\
    \hline
    Mean Q1 & $\mu \left\{\text{Q1 of column}\right\}$ & DD & .0789 \\
    \hline
    STD of Q1  & $\sigma \left\{\text{Q1 of column}\right\}$ & DD & .0613 \\
    \hline
    Mean Q3  & $\mu \left\{\text{Q3 of column}\right\}$ & DD & .0812 \\
    \hline
    STD of Q3  & $\sigma \left\{\text{Q3 of column}\right\}$ & DD & .1103 \\
    \hline
    Mean Percent of Z-Score Outliers  & 
    $\mu \left\{ \frac{\text{\# of Z-Score Outliers}}{|\mathbf{A}|_r} \right\}$ & ODP & .0558 \\
    \hline
    STD of Percent of Z-Score Outliers  &
    $\sigma \left\{ \frac{\text{\# of Z-Score Outliers}}{|\mathbf{A}|_r} \right\}$ & ODP & .0798 \\
    \hline
    Mean Correlation& $\mu \left\{\text{Correlation Matrix}\right\}$ & DL & .0559 \\
    \hline
    STD of Correlation & $\sigma \left\{\text{Correlation Matrix}\right\}$ & DL & .1526 \\
    \bottomrule
  \end{tabular}
  \captionsetup{font=footnotesize}
  \caption*{\textbf{Abbreviations}: DS: Data Size, DT: Data Type, DD: Data Distribution, ODP: Outlier Data Points, DL: Data Linearity, Weight: Variable Importance}
\end{table}

@@

Several observations were made from running these datasets. First, accuracy did not change for 4 of the 6 benchmarks. The other two benchmarks on eye states and diabetes prediction represented a loss of accuracy for SMEML by 0.13\% and 1.95\%, respectively. Training time, however, favored SMEML in every trial. SMEML outperformed the dummy framework by 173.96 seconds in its worst performance on the lung cancer trial and by 383.94 seconds in its best performance on the eye state dataset. Relative to the dummy's performance, a percent decrease in time to run for SMEML of at least 46.78\% could be seen, with a decrease of up to 69.22\% observerd as well. SMEML's worst performance by percent decrease was also the performance in which it returned perfect accuracy. 
\section{Discussion}
\subsection{Summary}
SMEML consistently achieves comparable accuracy to its dumb counterpart while significantly reducing the time required to train the model. Notably, SMEML's performance is not limited by the size of the dataset, achieving its best runtimes--first, third, and fourth fastest overall--on 'large' datasets (those exceeding 1000 rows). This efficiency is important in third world medicine where there is a constant influx of new patient data that must be processed quickly. A practical demonstration of this software's real-world application can be found on our SmartDx website [23]. Source code can be found at [24]. 
\subsection{Limitations}
Our implementation of SMEML has limitations. These include a small training set due to lack of available datasets, finite computing power, time constraints, and a lack of access to paticular scientific literature. Our framework only improves on the model selection process and does not address the various other aspects of AutoML. Therefore, SMEML is not comparable to other full-featured AutoML frameworks.
\subsection{Future Work}
Future work will include increasing the size of the training set, implementing a fully fledged AutoML framework, and conducting a detailed experiment comparing SMEML to other popular AutoML frameworks. Furthermore, we will examine other types of machine learning such as regression and multiclass-classification.

@@

[23] SmartDx Website. \url{https://smartdx.vercel.app/}.

[24] Github Repository. \url{https://github.com/ericspring08/SMEML-Paper}.
}

>>>>

Machine learning (ML) algorithms require complicated processes like hyperparamater tuning, and can be modified in conjunction with one another, like in ensemble learning. To do this on a larger scale and at greater speeds, recent research and effort have been put into developing AutoML frameworks. These frameworks create comfortable environments for models to excel in. They handle missing values, preprocess data, and optimize training to ensure the best results [1].

@@

When a user inputs a dataset into SMEML, it first extracts the attributes from the dataset and then feeds them into the SME to get the predicted rank of each model. Since we have determined that we can predict a top two model within the predicted top eight quite confidently, we can eliminate the bottom 20 models from the list. We run Bayseian optimization using scikit-optimize [16] on the top eight models to get the best hyperparamters for each model. We select the top models that have significantly better performance than other models. These models are stacked together and compared to the best individual model. The final model is the better one out of these two. This process is shown in Figure 1.
\begin{table}
  \caption{Attributes extracted from a dataset}
  \label{model-options-table}
  \centering
  \begin{tabular}{llll}
    \toprule
    Attribute & Formula & Purpose & Weight\\
    \midrule
    \# of Rows & $|\mathbf{A}|_r$ & DS & .0422 \\
    \hline
    \# of Columns & $|\mathbf{A}|_c$ & DS & .0415 \\
    \hline
    Target Distribution &
    $\frac{\max(\text{count}_1, \text{count}_0)}{|\mathbf{A}|_r}$
    & DT & .0489 \\
    \hline
    Numerical Columns &
    $\frac{|\mathbf{A}_{num}|}{|\mathbf{A}|_{c}}$
    & DT & .0428 \\
    \hline
    Binary Categorical Columns &
    $\frac{|\mathbf{A}_{bin}|}{|\mathbf{A}|_{c}}$
    & DT & .0117 \\
    \hline
    Mean \# of Distinct Values  & $\mu \left\{\text{\# of distinct values in column}\right\}$ & DT & .0076 \\ 
    \hline
    Mean IQR  & $\mu \left\{\text{IQR of column}\right\}$ & DD & .0594 \\ 
    \hline
    STD of IQR  & $\sigma \left\{\text{IQR of column}\right\}$ & DD & .0700 \\
    \hline
    Mean Q1 & $\mu \left\{\text{Q1 of column}\right\}$ & DD & .0789 \\
    \hline
    STD of Q1  & $\sigma \left\{\text{Q1 of column}\right\}$ & DD & .0613 \\
    \hline
    Mean Q3  & $\mu \left\{\text{Q3 of column}\right\}$ & DD & .0812 \\
    \hline
    STD of Q3  & $\sigma \left\{\text{Q3 of column}\right\}$ & DD & .1103 \\
    \hline
    Mean Percent of Z-Score Outliers  & 
    $\mu \left\{ \frac{\text{\# of Z-Score Outliers}}{|\mathbf{A}|_r} \right\}$ & ODP & .0558 \\
    \hline
    STD of Percent of Z-Score Outliers  &
    $\sigma \left\{ \frac{\text{\# of Z-Score Outliers}}{|\mathbf{A}|_r} \right\}$ & ODP & .0798 \\
    \hline
    Mean Correlation& $\mu \left\{\text{Correlation Matrix}\right\}$ & DL & .0559 \\
    \hline
    STD of Correlation & $\sigma \left\{\text{Correlation Matrix}\right\}$ & DL & .1526 \\
    \bottomrule
  \end{tabular}
  \captionsetup{font=footnotesize}
  \caption*{\textbf{Abbreviations}: DS: Data Size, DT: Data Type, DD: Data Distribution, ODP: Outlier Data Points, DL: Data Linearity, Weight: Variable Importance}
\end{table}

@@

Several observations were made from running these datasets. First, accuracy did not change for 4 of the 6 benchmarks. The other two benchmarks on eye states and diabetes prediction represented a loss of accuracy for SMEML by 0.13\% and 1.95\%, respectively. Training time, however, favored SMEML in every trial. SMEML outperformed the dummy framework by 173.96 seconds in its worst performance on the lung cancer trial and by 383.94 seconds in its best performance on the eye state dataset. Relative to the dummy's performance, a percent decrease in time to run for SMEML of at least 46.78\% could be seen, with a decrease of up to 69.22\% observerd as well. SMEML's worst performance by percent decrease was also the performance in which it returned perfect accuracy. 
\section{Discussion}
\subsection{Summary}
SMEML consistently achieves comparable accuracy to its dumb counterpart while significantly reducing the time required to train the model. Notably, SMEML's performance is not limited by the size of the dataset, achieving its best runtimes--first, third, and fourth fastest overall--on 'large' datasets (those exceeding 1000 rows). SMEML also chose several different kinds of algorithms rather than defaulting to industry-standard boosting algorithms [23]. This efficiency and versatility is important in third world medicine where there is a constant influx of new patient data that must be processed quickly. A practical demonstration of this software's real-world application can be found on our SmartDx website [24]. Source code can be found at [25].
\subsection{Limitations}
Our implementation of SMEML has limitations. These include a small training set due to lack of available datasets, finite computing power, time constraints, and a lack of access to paticular scientific literature. Our framework only improves on the model selection process and does not address the various other aspects of AutoML. Therefore, SMEML is not comparable to other full-featured AutoML frameworks.
\subsection{Future Work}
Future work will include increasing the size of the training set, implementing a fully fledged AutoML framework, and conducting a detailed experiment comparing SMEML to other popular AutoML frameworks. Furthermore, we will examine other types of machine learning such as regression and multiclass-classification.

@@

[23] Mayr, A. et al., "The Evolution of Boosting Algorithms From Machine Learning to Statistical Modelling." Methods Inf Med 2014; 53(06): 419-427. \url{https://doi.org/10.48550/arXiv.1403.1452}.

[24] SmartDx Website. \url{https://smartdx.vercel.app/}.

[25] Github Repository. \url{https://github.com/ericspring08/SMEML-Paper}.
}
